---
title: "Yoda1 and Mechanosome (Gilbert et al. 2024)"
output: html_notebook
---

### Load in data

```{r}
`%notin%` <- Negate(`%in%`)
mech <- read.csv(file.choose(), row.names = 1, stringsAsFactors = T)
yoda1 <- read.csv(file.choose(), row.names = 1, stringsAsFactors = T)
go<-read.csv(file.choose())
```

### Download and library packages

```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(version = "3.20")
BiocManager::install("clusterProfiler")
BiocManager::install("GO.db")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("DO.db")
install.packages("aPEAR")
BiocManager::install("rrvgo")
library(rrvgo)
library(aPEAR)
library(clusterProfiler)
library(GO.db)
library(org.Hs.eg.db)
library(DO.db)
library(readr)
```


### Process into correct format, create significant datasets

```{r}
mech$ensembl_ID <- gsub ("(.+)\\..+","\\1",rownames(mech))
mech$symbol <- mapIds(org.Hs.eg.db, mech$ensembl_ID, 'SYMBOL', 'ENSEMBL')
mechsig <- mech[mech$padj < 0.05 & !is.na(mech$padj),]
yoda1$symbol <- gsub ("(.+)\\..+","\\1",rownames(yoda1))
yoda1$ensembl_ID <- mapIds(org.Hs.eg.db, yoda1$symbol, 'ENSEMBL', 'SYMBOL')
yoda1sig <- yoda1[yoda1$padj < 0.05 & !is.na(yoda1$padj),]
```

### Split datasets into up and downregulated

```{r}
mech_up <- mech %>% filter(log2FoldChange > 0 & padj < 0.05) #4538
mech_down <- mech %>% filter(log2FoldChange < 0 & padj < 0.05) #3026
yoda1_up <- yoda1 %>% filter(log2FoldChange > 0 & padj < 0.05) #1601
yoda1_down <- yoda1 %>% filter(log2FoldChange < 0 & padj < 0.05) #1519
```

### Overall shared gene list

```{r}
yoda1_and_mech <- yoda1[yoda1$ensembl_ID %in% mech$ensembl_ID,] #9101
yoda1_mechsig <- yoda1sig[yoda1sig$ensembl_ID %in% mechsig$ensembl_ID,] #1476
```

### MA plot, GO enrichment, ontology, and aPEAR

```{r}
#MA plot
##Create datasets to be labelled
mech_biglog <- mech %>% filter(log2FoldChange < -2 | log2FoldChange > 2)
yoda1_biglog <- yoda1 %>% filter(log2FoldChange < -2 | log2FoldChange > 2)
yoda1mech_biglog <- yoda1_biglog[yoda1_biglog$ensembl_ID %in% mech_biglog$ensembl_ID,]
yoda1_and_mech$biglog <- match(yoda1_and_mech$symbol, yoda1mech_biglog$symbol)
yoda1_and_mech$plotnames_2 <- ifelse(yoda1_and_mech$biglog > 0, yoda1_and_mech$symbol, NA)
yoda1_ishlog <- yoda1 %>% filter(log2FoldChange < -1 | log2FoldChange > 1)
mech_ishlog <- mech %>% filter(log2FoldChange < -1 | log2FoldChange > 1)
yoda1mech_ishlog <- yoda1_ishlog[yoda1_ishlog$ensembl_ID %in% mech_ishlog$ensembl_ID,]
iv.mark <- yoda1_and_mech$biglog > 0
iv.sig <- yoda1_and_mech$padj < 0.05 
iv.up1 <- yoda1mech_ishlog$log2FoldChange > 1 & iv.sig
iv.dn1 <- yoda1mech_ishlog$log2FoldChange < -1 & iv.sig
iv.up <- yoda1_and_mechup$padj < 0.05
iv.dn <- yoda1_and_mechdown$padj < 0.05
##Make the plot
plot(log2(yoda1_and_mech$baseMean + 1), yoda1_and_mech$log2FoldChange, pch=".", col="grey",
     main="Yoda1 and Mechanosome", xlab="log2(baseMean)", ylab="Log2FC")
##Add labels
points(log2(yoda1mech_ishlog$baseMean + 1)[iv.up1], yoda1mech_ishlog$log2FoldChange[iv.up1], col="forestgreen", pch=20)
points(log2(yoda1mech_ishlog$baseMean + 1)[iv.dn1], yoda1mech_ishlog$log2FoldChange[iv.dn1], col="steelblue", pch=20)
points(log2(yoda1_and_mechup$baseMean + 1)[iv.up], yoda1_and_mechup$log2FoldChange[iv.up], col="forestgreen", pch=".")
points(log2(yoda1_and_mechdown$baseMean + 1)[iv.dn], yoda1_and_mechdown$log2FoldChange[iv.dn], col="steelblue", pch=".")
points(log2(yoda1_and_mech$baseMean + 1)[iv.mark], yoda1_and_mech$log2FoldChange[iv.mark], col=rgb(1,0.2,0.5,1), pch=20, cex = 2)
text(log2(yoda1_and_mech$baseMean + 1), yoda1_and_mech$log2FoldChange, yoda1_and_mech$plotnames_2, adj = c(0,0),
     cex = 0.5)

#GO enrichment
YODA1MECH_go <- enricher(yoda1_mechsig$ensembl_ID, universe = yoda1_and_mech$ensembl_ID, 
             TERM2GENE=go[,c(4,1)], TERM2NAME=go[,c(4,5)])
nrow(YODA1MECH_go@result[YODA1MECH_go@result$p.adjust < 0.05,]) #32
barplot(YODA1MECH_go, showCategory = 15, font.size = 7)
#create fullgo for use when applying ontologies
fullgo<-c(as.list(GOTERM),as.list(GOOBSOLETE))
YODA1MECH_go@result$Ontology<-as.factor(sapply(match(YODA1MECH_go@result$ID,names(fullgo)), 
                                     function(x) if (!is.na(x)) {Ontology(fullgo[[x]])} else {NA}))
#plot separate ontologies
mod.barplot(YODA1MECH_go, ont = "BP", title = paste("Biological Process Enrichment"), 
            showCategory = 15, font.size = 8)
mod.barplot(YODA1MECH_go, ont = "MF", title = paste("Molecular Function Enrichment"), 
            showCategory = 15, font.size = 8)
mod.barplot(YODA1MECH_go, ont = "CC", title = paste("Cellular Compartment Enrichment"), 
            showCategory = 15, font.size = 8)

#aPEAR
pear_yoda1mech <- findPathClusters(YODA1MECH_go@result, cluster = "hier" , minClusterSize = 10)
aPEAR::plotPathClusters(YODA1MECH_go@result, pear_yoda1mech$similarity, pear_yoda1mech$clusters)

#export dataset for further analysis
write.csv(yoda1_and_mech, file="C:/Users/hamay/OneDrive - Cardiff University/R Stuff/spreadsheets/yoda1_and_mech.csv")
```

### Shared upregulated genes

```{r}
#create shared upregulated dataset
yoda1_and_mechup <- yoda1_up[yoda1_up$ensembl_ID %in% mech_up$ensembl_ID,] #492
#run GO enrichment
YODA1MECHUP_go <- enricher(yoda1_and_mechup$ensembl_ID, universe = yoda1_and_mech$ensembl_ID, 
             TERM2GENE=go[,c(4,1)], TERM2NAME=go[,c(4,5)])
nrow(YODA1MECHUP_go@result[YODA1MECHUP_go@result$p.adjust < 0.05,]) #0
#add ontologies
YODA1MECHUP_go@result$Ontology<-as.factor(sapply(match(YODA1MECHUP_go@result$ID,names(fullgo)), 
                                     function(x) if (!is.na(x)) {Ontology(fullgo[[x]])} else {NA}))
#aPEAR clustering
pear_yoda1mechup <- findPathClusters(YODA1MECHUP_go@result, cluster = "hier" , minClusterSize = 20)
aPEAR::plotPathClusters(YODA1MECHUP_go@result, pear_yoda1mechup$similarity, pear_yoda1mechup$clusters)
#export for further analysis
write.csv(yoda1_and_mechup, file="C:/Users/hamay/OneDrive - Cardiff University/R Stuff/spreadsheets/yoda1_and_mechup.csv")
```

### Shared downregulated genes

```{r}
#create shared downregulated dataset
yoda1_and_mechdown <- yoda1_down[yoda1_down$ensembl_ID %in% mech_down$ensembl_ID,] #287
#run GO enrichment
YODA1MECHDN_go <- enricher(yoda1_and_mechdown$ensembl_ID, universe = yoda1_and_mech$ensembl_ID, 
             TERM2GENE=go[,c(4,1)], TERM2NAME=go[,c(4,5)])
nrow(YODA1MECHDN_go@result[YODA1MECHDN_go@result$p.adjust < 0.05,]) #1
barplot(YODA1MECHDN_go, showCategory = 15, font.size = 7)
#add ontologies
YODA1MECHDN_go@result$Ontology<-as.factor(sapply(match(YODA1MECHDN_go@result$ID,names(fullgo)), 
                                     function(x) if (!is.na(x)) {Ontology(fullgo[[x]])} else {NA}))
#aPEAR clustering
pear_yoda1mechdn <- findPathClusters(YODA1MECHDN_go@result, cluster = "hier" , minClusterSize = 18)
aPEAR::plotPathClusters(YODA1MECHDN_go@result, pear_yoda1mechdn$similarity, pear_yoda1mechdn$clusters)
#export data for further analysis
write.csv(yoda1_and_mechdown, file="C:/Users/hamay/OneDrive - Cardiff University/R Stuff/spreadsheets/yoda1_and_mechdown.csv")
```

### Mech up, yoda1 down

```{r}
yoda1dn_and_mechup <- yoda1_down[yoda1_down$ensembl_ID %in% mech_up$ensembl_ID,] #430
YODA1DMECHU_go <- enricher(yoda1dn_and_mechup$ensembl_ID, universe = yoda1_and_mech$ensembl_ID, 
             TERM2GENE=go[,c(4,1)], TERM2NAME=go[,c(4,5)])
nrow(YODA1DMECHU_go@result[YODA1DMECHU_go@result$p.adjust < 0.05,]) #15
barplot(YODA1DMECHU_go, showCategory = 15, font.size = 7)

YODA1DMECHU_go@result$Ontology<-as.factor(sapply(match(YODA1DMECHU_go@result$ID,names(fullgo)), 
                                     function(x) if (!is.na(x)) {Ontology(fullgo[[x]])} else {NA}))
mod.barplot(YODA1DMECHU_go, ont = "BP", title = paste("Biological Process Enrichment"), 
            showCategory = 15, font.size = 8)
mod.barplot(YODA1DMECHU_go, ont = "MF", title = paste("Molecular Function Enrichment"), 
            showCategory = 15, font.size = 8)
mod.barplot(YODA1DMECHU_go, ont = "CC", title = paste("Cellular Compartment Enrichment"), 
            showCategory = 15, font.size = 8)

pear_yoda1dmechu <- findPathClusters(YODA1DMECHU_go@result, cluster = "hier" , minClusterSize = 17)
aPEAR::plotPathClusters(YODA1DMECHU_go@result, pear_yoda1dmechu$similarity, pear_yoda1dmechu$clusters)
write.csv(yoda1dn_and_mechup, file="C:/Users/hamay/OneDrive - Cardiff University/R Stuff/spreadsheets/yoda1dn_and_mechup.csv")
```

### Yoda1 up, mech down

```{r}
yoda1up_and_mechdn <- yoda1_up[yoda1_up$ensembl_ID %in% mech_down$ensembl_ID,] #267
YODA1UMECHD_go <- enricher(yoda1up_and_mechdn$ensembl_ID, universe = yoda1_and_mech$ensembl_ID, 
             TERM2GENE=go[,c(4,1)], TERM2NAME=go[,c(4,5)])
nrow(YODA1UMECHD_go@result[YODA1UMECHD_go@result$p.adjust < 0.05,]) #30
barplot(YODA1UMECHD_go, showCategory = 15, font.size = 7)

YODA1UMECHD_go@result$Ontology<-as.factor(sapply(match(YODA1UMECHD_go@result$ID,names(fullgo)), 
                                     function(x) if (!is.na(x)) {Ontology(fullgo[[x]])} else {NA}))
mod.barplot(YODA1UMECHD_go, ont = "BP", title = paste("Biological Process Enrichment"), 
            showCategory = 15, font.size = 8)
mod.barplot(YODA1UMECHD_go, ont = "MF", title = paste("Molecular Function Enrichment"), 
            showCategory = 15, font.size = 8)

pear_yoda1umechd <- findPathClusters(YODA1UMECHD_go@result, cluster = "hier" , minClusterSize = 15)
aPEAR::plotPathClusters(YODA1UMECHD_go@result, pear_yoda1umechd$similarity, pear_yoda1umechd$clusters)
write.csv(yoda1up_and_mechdn, file="C:/Users/hamay/OneDrive - Cardiff University/R Stuff/spreadsheets/yoda1up_and_mechdn.csv")
```

### SubCellularVis (using gene symbols)
```{r}
devtools::install_github("jowatson2011/subcellularvis")
library(subcellularvis)
subcellularapp()
```
##### input all shared gene lists

### Revigo:

#### Yoda1 and mechanosome upregulated:

```{r}
simMatrix <- calculateSimMatrix(YODA1MECHUP_go@result$ID,
                                orgdb="org.Hs.eg.db",
                                method="Rel")
scores <- setNames(-log10(YODA1MECHUP_go@result$qvalue), YODA1MECHUP_go@result$ID)
reducedTerms <- reduceSimMatrix(simMatrix,
                                scores,
                                threshold=0.99999,
                                orgdb="org.Hs.eg.db")
scatterPlot(simMatrix, reducedTerms, algorithm = c("pca", "umap"), onlyParents = FALSE, size = "score", addLabel = TRUE, labelSize = 2)
```

#### Yoda1 and mechanosome downregulated:

```{r}
#Biological Process
simMatrixDn <- calculateSimMatrix(YODA1MECHDN_go@result$ID,
                                orgdb="org.Hs.eg.db",
                                method="Rel")
scoresdn <- setNames(-log10(YODA1MECHDN_go@result$qvalue), YODA1MECHDN_go@result$ID)
reducedTermsdn <- reduceSimMatrix(simMatrixDn,
                                scoresdn,
                                threshold=0.99999,
                                orgdb="org.Hs.eg.db")
scatterPlot(simMatrixDn, reducedTermsdn, algorithm = c("pca", "umap"), onlyParents = FALSE, size = "score", addLabel = TRUE, labelSize = 2)
```

GO terms/genes relating to mechanical load imported from mechanosome:

```{r}
mech_up_mechanosome <- read_table("C:/Users/hamay/OneDrive - Cardiff University/R Stuff/txt files/mechanically_up_mechanosome.txt") #46 genes
mech_down_mechanosome <- read_table("C:/Users/hamay/OneDrive - Cardiff University/R Stuff/txt files/mechanically_down_mechanosome.txt") #24
#find genes in yoda1 dataset
yoda1_and_mechgoup <- yoda1sig[yoda1sig$symbol %in% mech_up_mechanosome$Symbol,] #18
yoda1_and_mechgodn <- yoda1sig[yoda1sig$symbol %in% mech_down_mechanosome$Symbol,] #8
#specify regulation
yoda1up_and_mechgoup <- yoda1_up[yoda1_up$symbol %in% mech_up_mechanosome$Symbol,] #8
yoda1dn_and_mechgodn <- yoda1_down[yoda1_down$symbol %in% mech_down_mechanosome$Symbol,] #3
yoda1up_and_mechgoup
yoda1dn_and_mechgodn
```

Repeat for pain, bone remodelling, and inflammation

```{r}
#inflammation
inflam_up_mechanosome <- read_table("C:/Users/hamay/OneDrive - Cardiff University/R Stuff/txt files/inflam_up_mechanosome.txt") 
inflam_down_mechanosome <- read_table("C:/Users/hamay/OneDrive - Cardiff University/R Stuff/txt files/inflam_down_mechanosome.txt") 

yoda1_and_inflamgoup <- yoda1sig[yoda1sig$symbol %in% inflam_up_mechanosome$Symbol,] #22
yoda1_and_inflamgodn <- yoda1sig[yoda1sig$symbol %in% inflam_down_mechanosome$Symbol,] #9

yoda1up_and_inflamgoup <- yoda1_up[yoda1_up$symbol %in% inflam_up_mechanosome$Symbol,] #4
yoda1dn_and_inflamgodn <- yoda1_down[yoda1_down$symbol %in% inflam_down_mechanosome$Symbol,] #3
yoda1up_and_inflamgoup
yoda1dn_and_inflamgodn

#bone remodelling
bone_up_mechanosome <- read_table("C:/Users/hamay/OneDrive - Cardiff University/R Stuff/txt files/bone_up_mechanosome.txt") 
bone_down_mechanosome <- read_table("C:/Users/hamay/OneDrive - Cardiff University/R Stuff/txt files/bone_down_mechanosome.txt") 

yoda1_and_bonegoup <- yoda1sig[yoda1sig$symbol %in% bone_up_mechanosome$Symbol,] #19
yoda1_and_bonegodn <- yoda1sig[yoda1sig$symbol %in% bone_down_mechanosome$Symbol,] #13

yoda1up_and_bonegoup <- yoda1_up[yoda1_up$symbol %in% bone_up_mechanosome$Symbol,] #7
yoda1dn_and_bonegodn <- yoda1_down[yoda1_down$symbol %in% bone_down_mechanosome$Symbol,] #5
yoda1up_and_bonegoup
yoda1dn_and_bonegodn

#pain
pain_up_mechanosome <- read_table("C:/Users/hamay/OneDrive - Cardiff University/R Stuff/txt files/pain_up_mechanosome.txt") 
pain_down_mechanosome <- read_table("C:/Users/hamay/OneDrive - Cardiff University/R Stuff/txt files/pain_down_mechanosome.txt") 

yoda1_and_paingoup <- yoda1sig[yoda1sig$symbol %in% pain_up_mechanosome$Symbol,] #14
yoda1_and_paingodn <- yoda1sig[yoda1sig$symbol %in% pain_down_mechanosome$Symbol,] #7

yoda1up_and_paingoup <- yoda1_up[yoda1_up$symbol %in% pain_up_mechanosome$Symbol,] #8
yoda1dn_and_paingodn <- yoda1_down[yoda1_down$symbol %in% pain_down_mechanosome$Symbol,] #6
yoda1up_and_paingoup
yoda1dn_and_paingodn
```




