---
title: "Yoda1 analysis"
author: "Helena Mayfield"
date: "30/04/25"
output:
  html_notebook: 
    toc: true
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

Load in datasets

```{r}
go<-read.csv(file.choose())
yoda1 <- read.csv(file.choose(), row.names = 1, stringsAsFactors = T)
```

Download and install packages

```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(version = "3.20")
BiocManager::install("clusterProfiler")
BiocManager::install("GO.db")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("rrvgo")
install.packages("aPEAR")
library(aPEAR)
library(rrvgo)
library(clusterProfiler)
library(GO.db)
library(org.Hs.eg.db)
```

Format datasets and split into significantly regulated, upregulated, and downregulated by Yoda1 as opposed to controls

```{r}
yoda1$symbol <- gsub ("(.+)\\..+","\\1",rownames(yoda1))
yoda1$ensembl_ID <- mapIds(org.Hs.eg.db, yoda1$symbol, 'ENSEMBL', 'SYMBOL')
yoda1sig <- yoda1[yoda1$padj < 0.05 & !is.na(yoda1$padj),] #2848
yoda1_up <- yoda1 %>% filter(log2FoldChange > 0 & padj < 0.05) #1601
yoda1_down <- yoda1 %>% filter(log2FoldChange < 0 & padj < 0.05) #1519
```

Make an MA plot

```{r}
#Split groups to find genes with a fold change higher/lower than 2/-2
yoda1_up_2 <- yoda1_up %>% filter(log2FoldChange > 2) #1
yoda1_down_2 <- yoda1_down %>% filter(log2FoldChange < -2) #3
yoda1_biglog <- yoda1_sig %>% filter(log2FoldChange < -2 | log2FoldChange > 2)

#Format so that resulting genes can be labelled
yoda1$biglog <- match(yoda1$symbol, yoda1_biglog$symbol)
yoda1$plotnames_2 <- ifelse(yoda1$biglog > 0, yoda1$symbol, NA)

#Create values to aid in labelling
iv.mark <- yoda1$biglog > 0
iv.sig <- yoda1$padj < 0.05 
iv.up <- yoda1_ishlog$log2FoldChange > 1 & iv.sig
iv.dn <- yoda1_ishlog$log2FoldChange < -1 & iv.sig

#Create the MA plot
plot(log2(yoda1$baseMean + 1), yoda1$log2FoldChange, pch=".", col="grey",
    xlab="log2(baseMean)", ylab="Log2FC")
#Add labels
points(log2(yoda1_ishlog$baseMean + 1)[iv.up], yoda1_ishlog$log2FoldChange[iv.up], col="forestgreen", pch=20)
points(log2(yoda1_ishlog$baseMean + 1)[iv.dn], yoda1_ishlog$log2FoldChange[iv.dn], col="steelblue", pch=20)
points(log2(yoda1$baseMean + 1)[iv.mark], yoda1$log2FoldChange[iv.mark], col=rgb(1,0.2,0.5,1), pch=20, cex = 2)
text(log2(yoda1$baseMean + 1), yoda1$log2FoldChange, yoda1$plotnames_2, adj = c(0,0),
     cex = 0.5)
```

Enrich data for GO terms

```{r}
YODA1_go <- enricher(yoda1_sig$ensembl_ID, universe = yoda1$ensembl_ID, 
             TERM2GENE=go[,c(4,1)], TERM2NAME=go[,c(4,5)])
nrow(YODA1_go@result[YODA1_go@result$p.adjust <= 0.05,]) #139
barplot(YODA1_go, showCategory = 15, font.size = 7)

#create fullgo for use when applying ontologies
fullgo<-c(as.list(GOTERM),as.list(GOOBSOLETE))

#GO enrichment & plotting
YODA1_go@result$Ontology<-as.factor(sapply(match(YODA1_go@result$ID,names(fullgo)), 
                                     function(x) if (!is.na(x)) {Ontology(fullgo[[x]])} else {NA}))

#create function to plot separate ontologies
mod.barplot<-function(obj, ont, ...){
  obj@result<-obj@result[obj@result$Ontology==ont,]
  barplot(obj, ...)
}

#plot ontologies
mod.barplot(YODA1_go, ont = "BP", title = paste("Biological Process Enrichment"), 
            showCategory = 15, font.size = 7)

mod.barplot(YODA1_go, ont = "MF", title = paste("Molecular Function Enrichment"), 
            showCategory = 15, font.size = 7)

mod.barplot(YODA1_go, ont = "CC", title = paste("Cellular Compartment Enrichment"), 
            showCategory = 15, font.size = 7)

#rrvgo plot
simMatrixAll <- calculateSimMatrix(YODA1_go@result$ID,
                                orgdb="org.Hs.eg.db",
                                method="Rel")
scoresall <- setNames(-log10(YODA1_go@result$qvalue), YODA1_go@result$ID)
reducedTermsall <- reduceSimMatrix(simMatrixAll,
                                scoresall,
                                threshold=0.99999,
                                orgdb="org.Hs.eg.db")
scatterPlot(simMatrixAll, reducedTermsall, algorithm = c("pca", "umap"), onlyParents = FALSE, size = "score", addLabel = TRUE, labelSize = 2)
```

Create separate enrichments for up/downregulated groups

```{r}
#upregulated GO enrichment
YODA1UP_go <- enricher(yoda1_up$ensembl_ID, universe = yoda1$ensembl_ID, 
             TERM2GENE=go[,c(4,1)], TERM2NAME=go[,c(4,5)])
nrow(YODA1UP_go@result[YODA1UP_go@result$p.adjust <= 0.05,]) #51
barplot(YODA1UP_go, showCategory = 15, font.size = 7)
YODA1UP_go@result$Ontology<-as.factor(sapply(match(YODA1UP_go@result$ID,names(fullgo)), 
                                     function(x) if (!is.na(x)) {Ontology(fullgo[[x]])} else {NA}))
mod.barplot(YODA1UP_go, ont = "BP", title = paste("Biological Process Enrichment"), 
            showCategory = 15, font.size = 7)

mod.barplot(YODA1UP_go, ont = "MF", title = paste("Molecular Function Enrichment"), 
            showCategory = 15, font.size = 7)

mod.barplot(YODA1UP_go, ont = "CC", title = paste("Cellular Compartment Enrichment"),
            showCategory = 15, font.size = 7)

#upregulated rrvgo plot
#overall
simMatrixUp <- calculateSimMatrix(YODA1UP_go@result$ID,
                                orgdb="org.Hs.eg.db",
                                method="Rel")
scoresup <- setNames(-log10(YODA1UP_go@result$qvalue), YODA1UP_go@result$ID)
reducedTermsup <- reduceSimMatrix(simMatrixUp,
                                scoresup,
                                threshold=0.99999,
                                orgdb="org.Hs.eg.db")
scatterPlot(simMatrixUp, reducedTermsup, algorithm = c("pca", "umap"), onlyParents = FALSE, size = "score", addLabel = TRUE, labelSize = 3.8)
#downregulated GO enrichment
YODA1DOWN_go <- enricher(yoda1_down$ensembl_ID, universe = yoda1$ensembl_ID, 
             TERM2GENE=go[,c(4,1)], TERM2NAME=go[,c(4,5)])
write.csv(YODA1DOWN_go@result, file="C:/Users/hamay/OneDrive - Cardiff University/R Stuff/spreadsheets/yoda1downgo.csv")
nrow(YODA1DOWN_go@result[YODA1DOWN_go@result$p.adjust <= 0.05,]) #130
barplot(YODA1DOWN_go, showCategory = 15, font.size = 7)
YODA1DOWN_go@result$Ontology<-as.factor(sapply(match(YODA1DOWN_go@result$ID,names(fullgo)), 
                                     function(x) if (!is.na(x)) {Ontology(fullgo[[x]])} else {NA}))
mod.barplot(YODA1DOWN_go, ont = "BP", title = paste("Biological Process Enrichment"),
            showCategory = 15, font.size = 7)

mod.barplot(YODA1DOWN_go, ont = "MF", title = paste("Molecular Function Enrichment"),
            showCategory = 15, font.size = 7)

mod.barplot(YODA1_go, ont = "CC", title = paste("Cellular Compartment Enrichment"), 
            showCategory = 15, font.size = 7)
#downregulated rrvgo
#overall
simMatrixDn <- calculateSimMatrix(YODA1DOWN_go@result$ID,
                                orgdb="org.Hs.eg.db",
                                method="Rel")
scoresdn <- setNames(-log10(YODA1DOWN_go@result$qvalue), YODA1DOWN_go@result$ID)
reducedTermsdn <- reduceSimMatrix(simMatrixDn,
                                scoresdn,
                                threshold=0.99999,
                                orgdb="org.Hs.eg.db")
scatterPlot(simMatrixDn, reducedTermsdn, algorithm = c("pca", "umap"), onlyParents = FALSE, size = "score", addLabel = TRUE, labelSize = 3.8)
```

aPEAR for visualisation of gene sets

```{r}
#overall
pear_yoda1 <- findPathClusters(YODA1_go@result, cluster = 'hier', minClusterSize = 10)
aPEAR::plotPathClusters(YODA1_go@result, pear_yoda1$similarity, pear_yoda1$clusters)
#upregulated
pear_up <- findPathClusters(YODA1UP_go@result, cluster = "hier" , minClusterSize = 15)
aPEAR::plotPathClusters(YODA1UP_go@result, pear_up$similarity, pear_up$clusters)
#downregulated
pear_down <- findPathClusters(YODA1DOWN_go@result, cluster = "hier" , minClusterSize = 12)
aPEAR::plotPathClusters(YODA1DOWN_go@result, pear_down$similarity, pear_down$clusters)
```
NES = normalised enrichment score

Search for inflammation-related GO terms

```{r}
yoda1_go <- as.data.frame(YODA1_go)
yoda1_inflam <- yoda1_go %>% filter(ID == 'GO:0006954' | ID == 'GO:0002526' | ID == 'GO:0002544')
```
Only GO:0006954 is present. 60 of 2789 associated genes. Significantly upregulated

Search for bone remodelling-related GO terms
```{r}
yoda1_boneremod <- yoda1_go %>% filter(ID == 'GO:0001503' | ID == 'GO:0045779' | ID == 'GO:0045780' | ID == 'GO:0045453' | ID == 'GO:0030500' | ID == 'GO:0030501' | ID == 'GO:0030278' | ID == 'GO:0030279' | ID == 'GO:0043932' | ID == 'GO:0046849')
```
No matching GO terms

Search for pain-related GO terms
```{r}
yoda1_pain <- yoda1_go %>% filter(ID == 'GO:0019233' | ID == 'GO:0050966' | ID == 'GO:0048265' | ID == 'GO:1904057' | ID == 'GO:1904058' | ID == 'GO:0062149' | ID == 'GO:0051930' | ID == 'GO:0035255' | ID == 'GO:0035235' | ID == 'GO:0014047' | ID == 'GO:0005314' | ID == 'GO:0031798' | ID == 'GO:0031799' | ID == 'GO:0031800' | ID == 'GO:0031802')
```
No matching GO terms